A matrix-free approach for finite-strain hyperelastic problems using geometric multigrid

### 2018-11-02 doc/large_strain.tex
unify writing about material tangent

In order to solve \eqref{eq:weak_form} using Newton's method, a first order approximation around a given solution field $\overline{\gz u}$ is required such that
\begin{align}
F(\overline{\gz u} + \Delta \gz u, \delta \gz u) \approx
F(\overline{\gz u}, \delta \gz u) + \D_{\displaystyle \Delta \gz u} F(\overline{\gz u}, \delta \gz u) ,
\end{align}
where $\D_{\displaystyle \Delta \gz u}(\bullet)$ denotes the directional derivative in the direction $\Delta \gz u$.
For conservative traction boundary conditions, the directional derivative is given by
\begin{equation}
\begin{split}
\D_{\displaystyle \Delta \gz u} F(\overline{\gz u}, \delta \gz u)
&=
\int_{\mcl B_0}
\D_{\displaystyle \Delta \gz u} \left(\gz F \cdot \gz S \cdot \gz F^T\right)  :
\overline{\grad^s} \, \delta \gz u
\, \d V
\\
& +
\int_{\mcl B_0}
\overline{\gz \tau} :
\left[
  \Grad \, \delta \gz u \cdot
  \D_{\displaystyle \Delta \gz u} \gz F^{-1}
\right] \d V.
\end{split}
\label{eq:tangent_pre}
\end{equation}
Following \cite{Wriggers2008}, this can be simplified to\footnote{To that end
$\D_{\displaystyle \Delta \gz u} \gz F = \Grad \Delta \gz u$,
$\;\D_{\displaystyle \Delta \gz u} \gz F^{-1} = - \gz F^{-1} \cdot \D_{\displaystyle \Delta \gz u} \gz F \cdot \gz F^{-1}$ and
$\D_{\displaystyle \Delta \gz u} \gz E = \frac{1}{2}\left[\D_{\displaystyle \Delta \gz u} \gz F^T \cdot \gz F + \gz F^T \cdot \D_{\displaystyle \Delta \gz u} \gz F\right]$
are employed together with
$\D_{\displaystyle \Delta \gz u} \gz S = 2 \partial \gz S / \partial \gz C : \D_{\displaystyle \Delta \gz u} \gz E$.
}
\begin{equation}
  \begin{split}
\D_{\displaystyle \Delta \gz u} F(\overline{\gz u}, \delta \gz u)
  &=
  \int_{\mcl B_0} \overline{\grad^s} \Delta \gz u : J \boldsymbol{\mathcal{C}} : \overline{\grad^s} \, \delta \gz u \, \d V \\
  &+
  \int_{\mcl B_0}
  \overline{\grad}\delta \gz u :
  \left[
  \overline{\grad} \Delta \gz u \cdot
  \overline{\gz \tau}
  \right]
  \d V.
\end{split}
\label{eq:tangent}
\end{equation}
Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$, and the fourth-order {\color{red}spatial} tangent {\color{red}stiffness} tensor is the push forward of the material {\color{red}part of the referential} tangent {\color{red}stiffness tensor} $J \boldsymbol{\mathcal{C}} = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$.
The first term in \eqref{eq:tangent} is related to the linearization of the Piola-Kirchhoff stress $\gz S$ and {\color{red}is} therefore called material part of the tangent.
The second term in \eqref{eq:tangent} is called {\color{red}the} geometric part of the tangent since it originates from the linearization of $\gz F$, $\gz F^T$ and $\gz F^{-1}$.

@@

\begin{algorithm}[!ht]
  \SetKwInOut{Input}{Given}
  \SetKwInOut{Output}{Return}
  \Input{Source FE vector $\gz x$,
  cached $\gz \tau/J$ and $\boldsymbol{\mathcal{C}}$ for each cell and quadrature point,
  evaluated based on $\overline{\gz u^h}$}
  \Output{action of the FE tangent {\color{red}operator} \eqref{eq:algebraic_tangent} on $\gz x$}
  \ForEach{ element $K \in \Omega^h$ }{
        evaluate the following tensors at each quadrature point using sum factorization \;
        $\gz g_s := \overline{\grad^s} \gz x$ \tcp*{2nd order symmetric}
        $\gz g := \overline{\grad} \gz x$ \tcp*{2nd order}
        \ForEach{quadrature point $q$ on $K$}{
          evaluate $\boldsymbol{\mathcal{G}} \gz g := \gz g \cdot \left[\gz \tau/J\right]$ \tcp*{2nd order}
          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\overline{\grad} \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
          evaluate $\boldsymbol{\mathcal{C}}\gz g_s := \boldsymbol{\mathcal{C}}\gz : \gz g_s$ \tcp*{2nd order symmetric}
          queue $\boldsymbol{\mathcal{C}}\gz g_s$ for contraction $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;
        }
        evaluate queued contractions using sum factorization and distribute results to the destination vector
  }
  \caption{Matrix-free tangent operator: cache fourth-order material tangent {\color{red}stiffness tensor} $\boldsymbol{\mathcal{C}}$ and Kirchhoff stress $\gz \tau$.}
  \label{alg:mf_tensor4}
\end{algorithm}

@@

Figures \ref{fig:benchmark_miehe_Emmy} and \ref{fig:benchmark_miehe_IWR} show the results of the matrix-vector multiplication benchmarks as performed on the two clusters.
As expected, the matrix-vector multiplication becomes very expensive for higher order discretization for sparse matrix-based approaches.
The matrix-free implementation is faster than the matrix-based counterpart already for cubic Lagrangian basis in 2D and quadratic in 3D.
Figures \ref{fig:benchmark_miehe_Emmy_vmult2}, \ref{fig:benchmark_miehe_Emmy_vmult3}, \ref{fig:benchmark_miehe_IWR_vmult2} and \ref{fig:benchmark_miehe_IWR_vmult3} clearly show the influence of the different caching strategies.
Two conclusions can be drawn from these results:
The scalar caching implementation is the most time consuming of the three, as at each step we additionally evaluate the gradient of the displacement field in the {\color{red}referential} configuration that is required to evaluate the Kirchhoff stress $\gz \tau$.
There is little difference between the approach where the fourth-order material tangent {\color{red}stiffness} tensor is cached (``tensor4'') and the one where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the {\color{red}material part of the fourth-order spatial tangent stiffness tensor} on the second-order symmetric tensor.
This indicates that the time savings we get from the latter approach are insignificant within the complete matrix-free operator implementation.
Consequently, we can apply the matrix-free approach to any finite-strain material model by caching the fourth-order material tangent {\color{red}stiffness tensor} in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
It is therefore feasible to define highly performant generic operators that are independent of any applied constitutive laws.

>>>>

In order to solve \eqref{eq:weak_form} using Newton's method, a first order approximation around a given solution field $\overline{\gz u}$ is required such that
\begin{align}
F(\overline{\gz u} + \Delta \gz u, \delta \gz u) \approx
F(\overline{\gz u}, \delta \gz u) + \D_{\displaystyle \Delta \gz u} F(\overline{\gz u}, \delta \gz u) ,
\end{align}
where $\D_{\displaystyle \Delta \gz u}(\bullet)$ denotes the directional derivative in the direction $\Delta \gz u$.
For conservative traction boundary conditions, the directional derivative is given by
\begin{equation}
\begin{split}
\D_{\displaystyle \Delta \gz u} F(\overline{\gz u}, \delta \gz u)
&=
\int_{\mcl B_0}
\D_{\displaystyle \Delta \gz u} \left(\gz F \cdot \gz S \cdot \gz F^T\right)  :
\overline{\grad^s} \, \delta \gz u
\, \d V
\\
& +
\int_{\mcl B_0}
\overline{\gz \tau} :
\left[
  \Grad \, \delta \gz u \cdot
  \D_{\displaystyle \Delta \gz u} \gz F^{-1}
\right] \d V.
\end{split}
\label{eq:tangent_pre}
\end{equation}
Following \cite{Wriggers2008}, this can be simplified to\footnote{To that end
$\D_{\displaystyle \Delta \gz u} \gz F = \Grad \Delta \gz u$,
$\;\D_{\displaystyle \Delta \gz u} \gz F^{-1} = - \gz F^{-1} \cdot \D_{\displaystyle \Delta \gz u} \gz F \cdot \gz F^{-1}$ and
$\D_{\displaystyle \Delta \gz u} \gz E = \frac{1}{2}\left[\D_{\displaystyle \Delta \gz u} \gz F^T \cdot \gz F + \gz F^T \cdot \D_{\displaystyle \Delta \gz u} \gz F\right]$
are employed together with
$\D_{\displaystyle \Delta \gz u} \gz S = 2 \partial \gz S / \partial \gz C : \D_{\displaystyle \Delta \gz u} \gz E$.
}
\begin{equation}
  \begin{split}
\D_{\displaystyle \Delta \gz u} F(\overline{\gz u}, \delta \gz u)
  &=
  \int_{\mcl B_0} \overline{\grad^s} \Delta \gz u : J \boldsymbol{\mathcal{C}} : \overline{\grad^s} \, \delta \gz u \, \d V \\
  &+
  \int_{\mcl B_0}
  \overline{\grad}\delta \gz u :
  \left[
  \overline{\grad} \Delta \gz u \cdot
  \overline{\gz \tau}
  \right]
  \d V.
\end{split}
\label{eq:tangent}
\end{equation}
Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$, and the fourth-order {\color{red}material part of the spatial} tangent {\color{red}stiffness} tensor is the push forward of the material {\color{red}part of the referential tangent stiffness tensor} $J \boldsymbol{\mathcal{C}} = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$.
The first term in \eqref{eq:tangent} is related to the linearization of the Piola-Kirchhoff stress $\gz S$ and {\color{red}is} therefore called material part of the {\color{red}directional derivative}.
The second term in \eqref{eq:tangent} is called {\color{red}the} geometric part of the {\color{red}directional derivative} since it originates from the linearization of $\gz F$, $\gz F^T$ and $\gz F^{-1}$.

@@

\begin{algorithm}[!ht]
  \SetKwInOut{Input}{Given}
  \SetKwInOut{Output}{Return}
  \Input{Source FE vector $\gz x$,
  cached $\gz \tau/J$ and $\boldsymbol{\mathcal{C}}$ for each cell and quadrature point,
  evaluated based on $\overline{\gz u^h}$}
  \Output{action of the FE tangent {\color{red}operator} \eqref{eq:algebraic_tangent} on $\gz x$}
  \ForEach{ element $K \in \Omega^h$ }{
        evaluate the following tensors at each quadrature point using sum factorization \;
        $\gz g_s := \overline{\grad^s} \gz x$ \tcp*{2nd order symmetric}
        $\gz g := \overline{\grad} \gz x$ \tcp*{2nd order}
        \ForEach{quadrature point $q$ on $K$}{
          evaluate $\boldsymbol{\mathcal{G}} \gz g := \gz g \cdot \left[\gz \tau/J\right]$ \tcp*{2nd order}
          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\overline{\grad} \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
          evaluate $\boldsymbol{\mathcal{C}}\gz g_s := \boldsymbol{\mathcal{C}}\gz : \gz g_s$ \tcp*{2nd order symmetric}
          queue $\boldsymbol{\mathcal{C}}\gz g_s$ for contraction $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;
        }
        evaluate queued contractions using sum factorization and distribute results to the destination vector
  }
  \caption{Matrix-free tangent operator: cache {\color{red}material part of the fourth-order spatial tangent stiffness tensor} $\boldsymbol{\mathcal{C}}$ and Kirchhoff stress $\gz \tau$.}
  \label{alg:mf_tensor4}
\end{algorithm}

@@

Figures \ref{fig:benchmark_miehe_Emmy} and \ref{fig:benchmark_miehe_IWR} show the results of the matrix-vector multiplication benchmarks as performed on the two clusters.
As expected, the matrix-vector multiplication becomes very expensive for higher order discretization for sparse matrix-based approaches.
The matrix-free implementation is faster than the matrix-based counterpart already for cubic Lagrangian basis in 2D and quadratic in 3D.
Figures \ref{fig:benchmark_miehe_Emmy_vmult2}, \ref{fig:benchmark_miehe_Emmy_vmult3}, \ref{fig:benchmark_miehe_IWR_vmult2} and \ref{fig:benchmark_miehe_IWR_vmult3} clearly show the influence of the different caching strategies.
Two conclusions can be drawn from these results:
The scalar caching implementation is the most time consuming of the three, as at each step we additionally evaluate the gradient of the displacement field in the {\color{red}referential} configuration that is required to evaluate the Kirchhoff stress $\gz \tau$.
There is little difference between the approach where the {\color{red}material part of the fourth-order spatial tangent stiffness tensor} is cached (``tensor4'') and the one where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the {\color{red}material part of the fourth-order spatial tangent stiffness tensor} on the second-order symmetric tensor.
This indicates that the time savings we get from the latter approach are insignificant within the complete matrix-free operator implementation.
Consequently, we can apply the matrix-free approach to any finite-strain material model by caching the {\color{red}material part of the fourth-order spatial tangent stiffness tensor} in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
It is therefore feasible to define highly performant generic operators that are independent of any applied constitutive laws.
